---
title: "QSurvival do not group by censor status"
author: "Win-Vector LLC"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{QSurvival Censoring}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width=7)
```


```{r makedata,echo=FALSE,messages=FALSE,warnings=FALSE,results='hide'}
suppressWarnings(suppressMessages(library("ggplot2")))
suppressWarnings(suppressMessages(library('dplyr')))
suppressWarnings(suppressMessages(library('mgcv')))
set.seed(32535)

params <- list(groupA=list(k=1.8,lambda=50),
               groupB=list(k=0.8,lambda=10))
#plot(dweibull(1:100,shape=params[['groupA']][['k']],
#              scale=params[['groupA']][['lambda']]))
#plot(dweibull(1:100,shape=params[['groupB']][['k']],
#              scale=params[['groupB']][['lambda']]))

censorPoint = 50

mkData <- function(n) {
  d <- data.frame(group=sample(c('groupA','groupB'),n,replace=TRUE),
                  stringsAsFactors=FALSE)
  d$unobservedLifetime <- vapply(seq_len(nrow(d)),
                                 function(i) {
                                   max(1,round(rweibull(1,shape=params[[d[i,'group']]][['k']],
                                                        scale=params[[d[i,'group']]][['lambda']])))
                                 },
                                 numeric(1))
  d$observedLifetime <- pmin(d$unobservedLifetime,censorPoint)
  d$censored <- d$unobservedLifetime>d$observedLifetime
  d
}


dTrain <- mkData(200)
dTest <- mkData(101)

ggplot(dTrain,aes(x=unobservedLifetime,color=group)) +
  geom_density() + xlim(c(0,150)) + geom_vline(xintercept = censorPoint)
dTrain %>% dplyr::group_by(group) %>%
  dplyr::summarise(meanLifetime=mean(unobservedLifetime))

ggplot(dTrain,aes(x=observedLifetime,color=group)) +
  geom_density() 
dTrain %>% dplyr::group_by(group) %>%
  dplyr::summarise(meanLifetime=mean(observedLifetime))
```



```{r preparetomodel, echo=FALSE}
parallelCluster <- c()

## can't run paralle code in vignettes
#if(requireNamespace("parallel",quietly=TRUE)) {
#     parallelCluster <- parallel::makeCluster(4)
#}

ageLimit <- 45

prepTrainFull <- 
  QSurvival::buildQuasiObsForTraining(dTrain,
                                      dTrain$observedLifetime,
                                      ifelse(dTrain$censored,NA,dTrain$observedLifetime),
                                      'ID','observedAge','deathEvent',
                                      parallelCluster=parallelCluster)
# allow windowing further out than training without causing new levels.
# basically want time as shown to model (if you are showing time to mode)
# to end will before censoring region to make sure you are not extrapolating
# off a noisy end estimate.
prepTrainFull$surrogateAge <- pmin(prepTrainFull$observedAge,ageLimit)


obsWindow <- 200
prepTest <- 
  QSurvival::buildQuasiObsForComparison(dTest,
                                     obsWindow,
                                     dTest$observedLifetime,
                                     ifelse(dTest$censored,NA,dTest$observedLifetime),
                                     'ID','observedAge','deathEvent',
                                     parallelCluster=parallelCluster)
prepTest$surrogateAge <- pmin(prepTest$observedAge,45)


if(!is.null(parallelCluster)) {
  parallel::stopCluster(parallelCluster)
  parallelCluster <- NULL
}

```

*****

Reduced complexity of age portion of model, using spline over surrogate age.

```{r analyzegam}
# Could use s(observedAge), but using s(surrogateAge) means we don't use
# the GAM spline to extrapolate.
# Use degree 2 spline to approximate typical bathtub hazard shape.
model <- gam(deathEvent~group+s(surrogateAge,k=2),
             data=prepTrainFull,family=binomial)
print(summary(model))
prepTest$hazardPrediction <- as.numeric(predict(model,newdata=prepTest,type='response'))

testD <- QSurvival::summarizeHazard(prepTest,'ID','observedAge','hazardPrediction',
                           cumulativeHazardColumnName='cumulativeHazard',
                           survivalColumnName='survival',
                           deathIntensityColumnName='deathIntensity') 
testRes <- testD$details
```


*****

Bad estimates when grouping by censored, in particular notice
the uncensored group (those who died in the observation window)
have an inflated exepctedLifetime estimate (nearly 30 versus 17).
This is because estimate[lifetime] can be expanded as
(ignoring issues of non-independence of estimates);
(est_prob_censored) estimate[lifetime | censored] +
   (1-est_prob_censored) estimate[lifetime | not censored]
So if a non-censored example est_prob_censored is not near 0
they average in a significant extra lifetime.  Or in frequency terms
a non-negligible sub-population of the censored instance are inseprable
from the censored (long-lived) instances.

Roughly: think of the censoring mark as formally not observable just as 
y outcomes are considered not observable in standard regression.

```{r censoring}
# look at censored/uncensored issue
dTestAug <- dTest
dTestAug$ID <- seq_len(nrow(dTest))
dTestAug %>% dplyr::left_join(testD$expectedLifetime,by='ID') -> dTestAug
ggplot(data=dTestAug,aes(x=expectedLifetime,color=censored)) +
    geom_density(adjust=0.25)

# fairly good agreement when grouping by an observable variable
dTestAug %>% group_by(group) %>% summarize(
          expectedLifetime=mean(expectedLifetime),
          unobservedLifetime=mean(unobservedLifetime),
          observedLifetime=mean(observedLifetime))

# bad when grouping by censored indicator
dTestAug %>% group_by(censored) %>% summarize(
          expectedLifetime=mean(expectedLifetime),
          unobservedLifetime=mean(unobservedLifetime),
          observedLifetime=mean(observedLifetime))


```

Model on only non-censored data.  Notice we are now close (in expectation).

```{r noncensored}
prepTrainFullUncensored <- prepTrainFull[!prepTrainFull$censored,]
prepTestUncensored <- prepTest[!prepTest$censored,]
model <- gam(deathEvent~group+s(surrogateAge,k=2),
             data=prepTrainFullUncensored,family=binomial)
prepTestUncensored$hazardPrediction <- 
  as.numeric(predict(model,newdata=prepTestUncensored,type='response'))
testDU <- QSurvival::summarizeHazard(prepTestUncensored,'ID','observedAge','hazardPrediction',
                           cumulativeHazardColumnName='cumulativeHazard',
                           survivalColumnName='survival',
                           deathIntensityColumnName='deathIntensity') 
summary(testDU$expectedLifetime$expectedLifetime)
summary(prepTestUncensored$observedLifetime)
prepTestUncensored %>% group_by(group) %>% summarize(observedLifetime=mean(observedLifetime))
```

